---
title: "3. Development and evaluation of a gene expression-based EC index"
format:
  html:
    toc: true
    code-fold: true
    code-overflow: wrap
    code-tools: true
---

```{r setup}
knitr::opts_chunk$set(message = F,warning = F, eval = F)

# Colours
pals <- khroma::color('batlow')(8)
cols_trt <- pals[c(1, 6)]

suppressMessages(suppressWarnings({
  library(dplyr)
  library(ggplot2)
  library(patchwork)
  library(DESeq2)
  library(ggrepel)
  library(pheatmap)
  library(plotly)
  library(TCGAbiolinks)
  library(SummarizedExperiment)
  library(org.Hs.eg.db)
}))

```

# 1. Derive index on TCGA data

* Data are loaded from TCGA (primary tumor and solid normal tissue) and formatted appropriately.
  + EntrezIDs are converted to gene symbols and transcripts are collapsed.
* Resulting dataset is split 80%-20% into training and testing.

```{r loaddata, eval = F}
# Download TCGA expression data
p <- 'TCGA-UCEC'
query.exp <- GDCquery(project = p,
                      data.category = "Transcriptome Profiling",
                      data.type = "Gene Expression Quantification",
                      workflow.type = "STAR - Counts")
GDCdownload(query.exp, method = "api", files.per.chunk = 10)
data <- GDCprepare(query.exp)
save(data, file = "out/expr/tcga_expression.Rdata")

# keep only primary and solid normal tissue
data <- data[,data$definition!= 'Recurrent Solid Tumor']

# Visualise:
colData(data) |> as.data.frame() |> dplyr::group_by(patient) |> dplyr::mutate(n = n()) |> 
  dplyr::ungroup() |> 
  dplyr::filter(n!=1) |> 
  ggplot(aes(x = patient, y = n,
             fill = definition)) +
  geom_col()
# some individuals have more than one sample for tumour - unclear how to deal with best, -> exclude;
exclude <- colData(data) |> as.data.frame() |> dplyr::group_by(patient) |> dplyr::count() |> dplyr::filter(n >= 3) |> dplyr::pull(patient)
data <- data[,!data$patient %in% exclude]

# tr-test split (80-20):
ids <- unique(data$patient)
set.seed(2496)
ind <- sample(1:length(ids), length(ids)*0.2, replace = F)
split <- data.frame(patient = ids,
                    assign = rep("training", length(ids)))
split$assign[ind] <- 'testing'
ind <- match(data$patient, split$patient)
data$assignment <- split[ind,]$assign

# extract pheno and count data
pheno <- as.data.frame(colData(data))
cnt <- assays(data)$unstranded

# now we can collapse gene symbols
genes = mapIds(org.Hs.eg.db,
               keys=gsub("\\..*", "", rownames(cnt)), 
               column="SYMBOL",
               keytype="ENSEMBL",
               multiVals = 'first')
genedf <- data.frame(label = genes,
                     gene = names(genes)) |> 
  dplyr::distinct() |> dplyr::filter(!is.na(label))

cnt_collapse <- cnt |> as.data.frame() |> 
  tibble::rownames_to_column('gene') |> 
  dplyr::mutate(gene = gsub("\\..*", "", gene)) |> 
  dplyr::inner_join(genedf) |> 
  dplyr::group_by(label) |> 
  dplyr::summarise(across(where(is.numeric), sum)) |>
  dplyr::ungroup() |> 
  tibble::column_to_rownames('label')
# identical(colnames(cnt_collapse), rownames(pheno))

# Export tr / testing pheno and expression data
expTr <- cnt_collapse[,pheno$assignment == 'training']
phenoTr <- pheno[pheno$assignment=='training',]
expTe <- cnt_collapse[,pheno$assignment != 'training']
phenoTe <- pheno[pheno$assignment!='training',]
save(expTr, phenoTr, expTe, phenoTe, file = 'out/expr/trFiles.Rdata')


```

* Differential gene expression analysis is conducted on genes with ≥10 counts in more than 80% of the training set.

```{r deg, eval = F}
dds <- DESeqDataSetFromMatrix(countData = expTr,
                              colData = phenoTr,
                              design= ~ definition)
keep <- rowSums( counts(dds) >= 10 ) >= ncol(dds)*0.8
dds <- dds[keep,]
dds <- DESeq(dds)
res <- results(dds, name="definition_Solid.Tissue.Normal_vs_Primary.solid.Tumor")
save(dds, res, file = 'out/expr/deg.Rdata')
```

* Index training

```{r, eval = F}
sig <- as.data.frame(res[res$padj<0.01 & !is.na(res$padj),]) |> 
  tibble::rownames_to_column('gene')

# Get log counts for training and testing data
dds <- estimateSizeFactors(dds)
logcountsTr <- log2(counts(dds, normalized=TRUE) + 1 )

# same for testing data
dds_te <- DESeqDataSetFromMatrix(countData = expTe,
                              colData = phenoTe,
                              design= ~ definition)
dds_te <- estimateSizeFactors(dds_te)
logcountsTe <- log2(counts(dds_te, normalized=TRUE) + 1 )

source("~/Dropbox/src/DNAm-classifiers/el-classifier.R")
phenoTr$type <- ifelse(phenoTr$definition=='Solid Tissue Normal', 0, 1)
phenoTe$type <- ifelse(phenoTe$definition=='Solid Tissue Normal', 0, 1)

res <- el_classifier(beta_tr = logcountsTr[sig$gene,],
                   type_tr = phenoTr$type,
                   beta_val = logcountsTr[sig$gene,],
                   type_val = phenoTr$type,
                   alpha = 0.5)
# res$oob_auc
# res$val_roc
weights <- as.matrix(coef(res$fit.cv))
weights <- weights[weights != 0,]
weights <- weights[names(weights) != '(Intercept)']
save(weights, file = 'out/exp-index/exp-index.Rdata')

B <- logcountsTr[names(weights),]
scale <- apply(B, 2, function(x) sum(x*weights))
save(scale, file = 'out/exp-index/exp-index-scale.Rdata')

load('out/exp-index/exp-index.Rdata')
genes <- as.data.frame(weights) |> tibble::rownames_to_column('gene') |> dplyr::rename(coefficient = weights)
write.table(genes, file = 'out/GeneSupp.csv',
            sep = ',', row.names = F)

```

* Volcano plot with index genes highlighted

```{r degplots, eval = F}
load('out/expr/deg.Rdata')
tmp <-  res |> as.data.frame() |>
  tibble::rownames_to_column('gene') |> 
  dplyr::mutate(sig = case_when(padj < 0.01 & log2FoldChange < 0 ~ 'down',
                                padj < 0.01 & log2FoldChange > 0 ~ 'up',
                                TRUE ~ NA)) 


volcano <- tmp |> 
  dplyr::mutate(lab = ifelse(-log10(pvalue) > 60, gene, ''),
                label = ifelse(gene %in% names(weights), gene, '')) |> 
  ggplot(aes(x = log2FoldChange,
             y = -log10(pvalue),
             colour = -log10(pvalue))) +
  geom_point(aes(size = -log10(pvalue)),
             alpha = 0.3) +
  scale_colour_viridis_c(end = 0.85) +
  theme_bw() +
  scale_size_continuous(range = c(0.2, 1.5),
                        guide = F) +
  geom_text_repel(aes(label = label),
                  size = 2.5,
                  max.overlaps = 50)

save(volcano, file = 'out/geneVolcano.Rdata')
ggsave(volcano, file = 'out/plots/expVolcano.pdf',
       width = 5.5, height = 5, units = 'in',
       dpi = 600)

# pvalHist <- tmp |> 
#   ggplot(aes(x = pvalue)) +
#   geom_histogram(bins = 200) +
#   theme_bw() +
#   labs(x = 'p value', y = 'n')
```

* Individual genes and heatmap:

```{r, eval = F}
# Sanity plots of individual genes
tmp <- cbind(phenoTr, t(logcountsTr[names(weights),])) |> 
  tidyr::pivot_longer(names(weights),
                      names_to = 'gene',
                      values_to = 'value')
tmp |> 
  ggplot(aes(x = as.factor(type),
             y = value,
             fill = as.factor(type))) +
  geom_boxplot(outlier.shape = NA) +
  facet_wrap(~gene, scales = 'free_y') +
  theme_bw() +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        legend.position = 'top') +
  scale_fill_manual(values = pals[c(1, 5)],
                    labels = c("Control", "Case"),
                    name = '') +
  labs(x = '', y='log2(count+1)')

tmp <- cbind(phenoTe, t(logcountsTe[names(weights),])) |> 
  tidyr::pivot_longer(names(weights),
                      names_to = 'gene',
                      values_to = 'value')
tmp |> 
  ggplot(aes(x = as.factor(type),
             y = value,
             fill = as.factor(type))) +
  geom_boxplot(outlier.shape = NA) +
  facet_wrap(~gene, scales = 'free_y') +
  theme_bw() +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        legend.position = 'top') +
  scale_fill_manual(values = pals[c(1, 5)],
                    labels = c("Control", "Case"),
                    name = '') +
  labs(x = '', y='log2(count+1)')

# Heatmap
library(ComplexHeatmap)
mat <- logcountsTr[names(weights),]
ms <- scale(mat)
# identical(colnames(ms), phenoTr$barcode)
anno <- columnAnnotation(type = ifelse(phenoTr$type == 0, 'Normal', 'Cancer'),
                         col = list(type = c('Normal' = pals[1], 'Cancer' = pals[6])))
# anno <- columnAnnotation(type = ifelse(phenoTr$tissue_type == 'Normal', 'Normal', 'Cancer'),
#                          col = list(type = c('Normal' = pals[1], 'Cancer' = pals[6])))
hm <- Heatmap(ms,
        show_row_names = T,
        show_column_names = F,
        name = 'expression\nz-scaled',
        row_names_gp = gpar(fontsize = 6),
        top_annotation = anno)

save(hm, file = 'out/geneEx_heatmap.Rdata')
pdf(file = 'out/plots/expHeatmap.pdf', width = 6.5, height = 5.75)
print(hm)
dev.off()

```

Compute final index in TCGA

```{r}
ECindex.exp <- function(exp){
  load('out/exp-index/exp-index-scale.Rdata')
  load('out/exp-index/exp-index.Rdata')
  intersect <- intersect(names(weights), rownames(exp))
  intersect <- weights[intersect]
  B <- exp[names(intersect),]
  tmp <- apply(B, 2, function(x) sum(x*intersect))
  out <- (tmp - mean(scale))/sd(scale)
  
  return(out)
}

outTr <- ECindex.exp(logcountsTr)
outTe <- ECindex.exp(logcountsTe)

tmpTr <- phenoTr
# identical(phenoTr$barcode, names(outTr))
tmpTr$index <- outTr

tmpTe <- phenoTe
# identical(phenoTe$barcode, names(outTe))
tmpTe$index <- outTe

data_tmp <- rbind(tmpTr, tmpTe) |> 
  dplyr::mutate(set = assignment,
                type = ifelse(tissue_type == 'Normal', 'Normal', 'Cancer')) |> 
                # type = ifelse(type == 0, 'Normal', 'Cancer')) |> 
  dplyr::select(index, barcode, type, set)
```

External validation: GEO Data -> download and compute

```{r}
# Apply in GEO data
gse <- getGEO("GSE63678")
geo_exp <- exprs(gse$GSE63678_series_matrix.txt.gz)
# BiocManager::install("hgu133plus2.db")
# BiocManager::install("affycoretools")
library(hgu133plus2.db)
library(affycoretools)

eset <- ExpressionSet(assayData=geo_exp)
eset <- annotateEset(eset, hgu133plus2.db)
meta <- as.data.frame(featureData(eset)@data)

ind <- match(rownames(geo_exp), meta$PROBEID)
rownames(geo_exp) <- meta$SYMBOL

pheno <- as.data.frame(phenoData(gse$GSE63678_series_matrix.txt.gz)@data)
# identical(colnames(geo_exp), rownames(pheno))

# sum(names(weights) %in% rownames(geo_exp))
pheno$index <- ECindex.exp(geo_exp)

pheno <- pheno |> 
  dplyr::filter(`tissue:ch1` == 'endometrium') |> 
  dplyr::mutate(set = 'external validation',
                type = ifelse(`disease state:ch1` == 'normal', 'Normal', 'Cancer'),
                barcode = NA) |> 
  dplyr::select(index, set, barcode, type)
save(pheno, file = 'out/pheno_expExtVal.Rdata')
```

Combine and plot:

```{r}
# load('out/pheno_expExtVal.Rdata')

# data_exprtest <- rbind(data_tmp, pheno)|>
#   dplyr::mutate(type = factor(type, levels = c("Normal", "Cancer")),
#                 set = factor(set, levels = c('training', 'testing', 'external validation')))

data_exprtest <- data_tmp |>
  dplyr::mutate(type = factor(type, levels = c("Normal", "Cancer")),
                set = factor(set, levels = c('training', 'testing')))

tcga_ec_exp_boxes <- data_exprtest |>
  ggplot(aes(x = type,
             y = index)) +
  geom_boxplot(aes(fill = type),
               alpha = 0.2) +
  ggbeeswarm::geom_quasirandom(aes(colour = type),
                               size = 0.9) +
  theme_bw() +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        legend.position = 'top') +
  facet_wrap(~set) +
  scale_fill_manual(values = pals[c(1, 6)],
                    labels = c("Normal", "Cancer"),
                    name = '',
                    aesthetics = c('colour', 'fill')) +
  labs(x = '', y='TCGA-EC expression index')

save(tcga_ec_exp_boxes, file = 'out/exp_index_development.Rdata')
ggsave(tcga_ec_exp_boxes, file = 'out/plots/expBoxes.pdf',
       width = 8, height = 3.5, units = 'in',
       dpi = 600)

# AUC plots:
source("~/Dropbox/src/plot/plot_roc_slim.R")

tcga_ec_exp_auc_tr <- plot_roc(data_exprtest[data_exprtest$set == 'training',]$type,
              data_exprtest[data_exprtest$set == 'training',]$index,
              size = 2.8
              ) +theme(aspect.ratio = NULL)
tcga_ec_exp_auc_val <- plot_roc(data_exprtest[data_exprtest$set == 'testing',]$type,
              data_exprtest[data_exprtest$set == 'testing',]$index,
              size = 2.8
              ) +theme(aspect.ratio = NULL)
# tcga_ec_exp_auc_extval <- plot_roc(data_exprtest[data_exprtest$set == 'external validation',]$type,
#               data_exprtest[data_exprtest$set == 'external validation',]$index,
#               size = 2.8
#               ) +theme(aspect.ratio = NULL)

save(tcga_ec_exp_auc_tr, tcga_ec_exp_auc_val, file = 'out/expAUCs_plots.Rdata')
aucs <- (tcga_ec_exp_auc_tr|tcga_ec_exp_auc_val)

ggsave(aucs, file = 'out/plots/expAucs.pdf',
       width = 8, height = 3.5, units = 'in',
       dpi = 600)
```

# Joint figure on training AUCs with DNAme

```{r}
load('out/DNAmAUCs_plots.Rdata')
load('out/expAUCs_plots.Rdata')

load('out/geneVolcano.Rdata')
load('out/geneEx_heatmap.Rdata')

tcga_ec_auc_tr + tcga_ec_auc_val + tidyHeatmap::wrap_heatmap(hm)
```

assemble manually because heatmap otherwise can't be included.